### 2014-12-01 paper/intro.tex
examples

>>>>

\subsection*{Unique Pointers}
In Rust, unique pointers are the owners of heap memory.
Heap memory is allocated when a new unique pointer is created.
Heap memory is freed when a unique pointer falls out of scope.
\begin{verbatim}
{
  let x: ~int; // A stack variable containing a unique pointer to an integer
  x = ~3; // Allocates a heap integer, initializes it with 3, and stores the pointer in x
  // The heap memory owned by x is freed when it falls out of scope
}
\end{verbatim}

To avoid double frees, unique pointers must remain unique.
Thus, when a unqiue pointer would be copied the original must be rendered unsuable.
That is, they are moved rather than copied.
\begin{verbatim}
{
  let x: ~int = ~3;
  let y: ~int = x; // x is moved into y. x is no longer usable.
  let z: ~int = x; // ERROR!
}
\end{verbatim}

However, these deinitialized paths can be reinitialized.
\begin{verbatim}
{
  let x: ~int = ~3;
  let y: ~int = x; // x is now deinitialized
  x = ~1; // x is initialized again
}
\end{verbatim}

Unique pointers can also be freed if they would be orphaned by assignment.
\begin{verbatim}
{
  let x: ~~int = ~~3;
  *x = ~1; // The ~3 that *x points too would be orphaned here. It is freed instead.
}
\end{verbatim}

This is not a dynamic check. 
The compiler detects when heap memory should be freed and inserts the necessary code.
To check that the Rust compiler does this correctly, Patina uses a shallow free statement.
The only way to deallocate a unique pointer in Patina is with these explicit frees.
\begin{verbatim}
{
  let x: ~~int = ~~3;
  //free(x); // ERROR! would orphan *x
  // The proper way
  free(*x);
  free(x);
}
\end{verbatim}
However, the free statement is only valid for initialized pointers, which prevents double frees.
\begin{verbatim}
{
  let x: ~int = ~3;
  free(x); // *x is now deallocated. x is uninitialized
  free(x); // ERROR! x is not initialized
}
\end{verbatim}
Finally, the free statement requires that the pointer is unaliased, preventing dangling pointers.
\begin{verbatim}
{
  let x: ~int = ~3;
  let y: &int = &*x; // Create an immutable borrowed reference to *x
  free(x); // ERROR! would make y a dangling pointer
}
\end{verbatim}

\subsection*{Borrowed References}
Borrowed references are a means to provide safe, temporary aliases in Rust.
They do so by enforcing two invariants.
The first is that references are always valid.
%They do so by enforcing two invariants: 
%references are always valid and
%no memory is ever simultaneously aliased and mutable.
%As with unique pointers, these are static checks.
\begin{verbatim}
{
  let x: int; // An uninitialized stack integer
  let y: &int = &x; // ERROR! *y would point to uninitialized memory
}
\end{verbatim}
This also extends to cases where the referent falls out of scope too soon.
\begin{verbatim}
{
  let y: &int;
  {
    let x: int = 3;
    y = &x; // ERROR! x does not live long enough
  } // x would be freed here
  // *y would point to unallocated memory here
}
\end{verbatim}
The second invariant is to prevent any memory from being simultaneously aliased and mutable.
This is to avoid situtations like the following:
\begin{verbatim}
{
  let x: Option<int> = Some(3);
  let y: &int = match x {
    Some(ref z) => z,
    None => fail!(), // Impossible
  }
  // If x is mutable, then the payload 3 is both aliased and mutable
  x = None // The payload is now uninitialized so *y is a dangling pointer
}
\end{verbatim}
Therefore immutable borrows freeze the underlying data.
It cannot be mutated by anything while the borrow exists.
However, it is freely copyable.
\begin{verbatim}
{
  let x: int = 3;
  let y: &int = &x;
  let z: &int = y; // Both y and z are readable
  x = 2; // ERROR! x is frozen by y
}
\end{verbatim}
Conversely, mutable borrows forbid all other access to the underlying data.
It does allow the data to be mutated, but it is an affine datatype like unique pointers.
\begin{verbatim}
{
  let x: int = 3;
  let y: &mut int = &mut x;
  //let z: int = x; // ERROR! cannot use x while mutable borrowed
  let z: int = *y; // OK

  let w: &mut int = y; // Moves y into w. y is not uninitialized
  z = *y; // ERROR! y is uninitialized
}
\end{verbatim}
As with unique pointers, these checks are all static.

### 2014-12-01 paper/paths.tex
examples

\section*{Runtime Typing}
\subsection*{Route Typing}
Typing routes is much like typing paths; however, the $\pay{r}$ route requires
special attention. Since this route points to the payload of a variant,
its type depends upon the value of the discriminant of the variant.
We handle this with an additional context that records the discriminant value
of a variant stored at a route.

@@

@@

\section*{Shadow Coherence}
Of course, we also need to check that our shadow heap correctly models
the actual heap. For the most part, it is exactly the relation you would expect.
%We require two auxiliary judgments to check for certain initialization states of layouts.
%The $\textsc{init}(H,l)$ judgment checks that $l$ is \emph{fully initialized}, 
%i.e. $\void$ does not occur in any memory reachable from $l$.
%These layouts are safe to use because we cannot access uninitialized memory from them.
The use of \textsc{init} in \textsc{LS-REF} and \textsc{LS-VAR} ensures that
the pointees of borrowed references and the payload of initialized variants
are fully initialized.
The $\textsc{droppable}(l)$ judgment checks that $l$ owns no initialized heap memory,
i.e. $\own{\addr{\ga}{r}}$ is not reachable from $l$ except through references.
These layouts are safe to deallocate because we cannot orphan heap memory by doing so.
\newline

>>>>

\section*{Runtime Typing}
\subsection*{Route Typing}
Typing routes is much like typing paths; however, the $\pay{r}$ route requires
special attention. Since this route points to the payload of a variant,
its type depends upon the value of the discriminant of the variant.
This is used for handling match by reference.
The match arms have a pointer to the payload in scope, but the type of the
payload depends upon the value of the discrimiant.
For example,
\begin{verbatim}
  // x: <[], int> is a Patina encoding of Option<int>
  match x by imm y // y will have route `pay r' were `r' is the route of x
    // [] (None) case
    {
      // y: & imm [] is in scope here
    }
    // int (Some) case
    {
      // y: & imm int is in scope here
    }
\end{verbatim}

We handle this with an additional context that records the discriminant value
of a variant stored at a route.

@@

Two examples are useful for understanding shadows.
First, consider nested pointers:
\begin{verbatim}
{
  let x: ~~int = ~~3;
  // Here the shadow of x is `~ ~ int'
  // x, *x, and **x are all initialized

  let y: ~int = *x; // *x is now deinitialized
  // Here the shadow of x is `~ uninit' and the shadow of y is `~ int'
  // x is initialized, by *x and **x are not
  // both y and *y are initialized
}
\end{verbatim}

Second, consider tuples:
\begin{verbatim}
{
  let x: [~int, ~int] = [~1, ~2];
  // Here the shadow of x is [~int, ~int]
  // x, x.1, *(x.1), x.2, and *(x.2) are all initialized

  free(x.1);
  // Here the shadow of x is [uninit, ~int]
  // x, x.1, x.2, and *(x.2) are all initialized, but *(x.1) is not
}
\end{verbatim}

The shadow heap gives the typechecker enough information 
to ensure we do not use or dereference uninitialized memory.
For example, here is how it can catch dereferencing a null pointer:
\begin{verbatim}
{
  let x: ~~int = ~~3;
  free(*x); // deallocates **x, leaving *x uninitialized
  // at this point the shadow of x is `~ uninit'
  let y: int = **x; // ERROR! dereferences a null pointer *x
  // the type checker know the shadow of *x is `uninit'
  // if a further dereference was allowed (i.e. **x), a null pointer derefernce would occur
  // the type checker can avoid null pointer dereferences by forbidding derefs of `uninit' data

}
\end{verbatim}

It can also catch a missing free that would orphan memory.
\begin{verbatim}
{
  let x: ~int = ~3;
  // the shadow of x is `~ int'

  // the stack variable x will be popped off the stack, orphaning the heap memory at *x
  // the type checker can identify this error by seeing that the shadow of x contains `~' data
  // by requiring that shadows cannot contain `~' when freed, the type checker prevents leaks
}
\end{verbatim}

@@

\section*{Shadow Coherence}
Of course, we also need to check that our shadow heap correctly models
the actual heap. For the most part, it is exactly the relation you would expect.
%We require two auxiliary judgments to check for certain initialization states of layouts.
%The $\textsc{init}(H,l)$ judgment checks that $l$ is \emph{fully initialized}, 
%i.e. $\void$ does not occur in any memory reachable from $l$.
%These layouts are safe to use because we cannot access uninitialized memory from them.
The use of \textsc{init} in \textsc{LS-REF} and \textsc{LS-VAR} ensures that
the pointees of borrowed references and the payload of initialized variants
are fully initialized.
The $\textsc{droppable}(l)$ judgment checks that $l$ owns no initialized heap memory,
i.e. $\own{\addr{\ga}{r}}$ is not reachable from $l$ except through references.
These layouts are safe to deallocate because we cannot orphan heap memory by doing so.
The $\tc{V}{H}{\gU}$ judgment ensures that the shadow heap corresponds to the real heap.
\newline
